A family of accessibility measures derived from spatial interaction principles

### 2025-02-21 paper/manuscript.qmd
cleaned up draft up until singly-constrained access. Consolidated sections, removed some commented-out work, decided against including an emp. example for now.

---
title: "A family of accessibility measures derived from spatial interaction principles"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
includes:
      in_header: 'gt_packages.sty'
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = '', 
  out.width = "1\\linewidth")
```

@@

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished. There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships when discussing population, opportunities, and their separation in space At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".

@@

Parting from this, it is possible to derive a family of spatial interaction models. In Wilson's terms, there is the "unconstrained" (but actually total flow constrained) model in @eq-phys-gravity-model. The proportionality constant $k$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:

@@

The work of @hansen1959 and @wilson1971 responded to important developments in the Anglo-American world at the time, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994]. Said dictates and needs were far from trivial. In the US alone, the Federal-Aid Highway Act of 1956 authorized the creation of the US Interstate Highway System, with a budget of more than one hundred billion dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007]. Spatial interaction modelling, with its ability to quantify trips, was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016]. Accessibility at the time did not quite have that power, as it did not quantify trips, but rather something somewhat more elusive: the less tangible "potential for interaction". In this way, where spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.

@@

After compiling this corpus of documents, we used the {bibliometrix} package [@bibliometrix2017] to create a bibliographical coupling matrix. In this matrix, two documents are coupled if they share at least one common reference, and the strength of the coupling increases with the number of references that the documents have in common. @fig-Fig1 presents the results. In this figure, each symbol represents a document and their distance on the plot represents the strength of bibliographical coupling: more strongly coupled documents (i.e., those sharing more items in their lists of references) are plotted closer to each other.

@@

```{r}
#| label: fig-Fig1
#| out-width: 70%
#| fig-cap: "Bibliometric coupling of papers that cite Hansen (1959) and/or Wilson (1971)."

@@

@@

Using these constraints, the SIM @eq-phys-gravity-model is distinguished into three general cases. These cases are presented in increasing order of constraint restrictiveness and continued to be useful in classifying the SIM in trip distribution modelling to this day [@ortuzar_2011_modelling].
-->

@@

As we argued in the preceding section, the accessibility and spatial interaction modelling literature streams tended to evolve with relatively little contact after @hansen1959 and @wilson1971. This may explain the failure of the constraints/proportionality constant(s) of spatial interaction models to cross over to accessibility analysis. This is intriguing, considering that Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the similarity between the denominator of @eq-production-constrained-balancing-factor and Hansen's accessibility indicator. In fact, Wilson noted that Hansen's accessibility was the inverse of $A_i$, which offered a potential interpretation for the proportionality constants $A_i$ (p. 10):

@@

This, however, only helps to illustrate the ease with which confusion may arise in this context. The relationship in @eq-Ai-as-accessibility appears to be misguided when we note that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force:

$$
F = G\frac{M_1\cdot M_2}{d_{12}^2}
$$

\noindent to the pairwise population potential:
$$
V_1 = G\frac{M_2}{d_{12}}
$$

\noindent and from there to the system-wide population potential (after losing $G$):
$$
V_i = \sum_j \frac{M_j}{d_{ij}}
$$

It is useful to think of the accessibility as a partial sum of the spatial interaction, before we return to Wilson's general model:

$$
T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
$$

Then, a way to define the _potential for interaction_ (i.e., accessibility) is as follows:

@@

<!--
. Potential interaction is the preoccupation of accessibility measures. The same analytical framework to define the balancing factors $k$ can be used for the family of accessibility measures as in the family of SIMs, however with different interpretation. In this section, we explicitly define these cases. 
-->

Next, we show how various accessibility measures can be defined depending on how we choose to constrain them.

@@

Unconstrained accessibility results when we set the proportionality constant/balancing factor to one in @eq-access-01:

@@

Then, the partial sum becomes:

@@

We can see that in this case we simply have Hansen's accessibility. In general, the following does not hold true:

$$
\sum_i V^0_{i} = O
$${#eq-no-constraint0-accessibility}

\noindent where $O$ is the total number of opportunities in the system. In other words, the sum of all accessibility scores will generally be different from the total of the opportunities available in the region. To complicate matters, the difference will depend on the choice of impedance function $f(c_{ij})$ and any associated parameters, as well as the number of locations $n$ for which an accessibility score is calculated. For this reason, unconstrained accessibility can only be appropriately used as an ordinal variable to make comparisons of size (greater than, less than, equal to), but not to calculate intervals (the magnitude of differences) or ratios.

## Total opportunity constrained accessibility

We can choose to retain the proportionality constant in @eq-access-01 to obtain the following:

@@

Our accessibility measure now becomes:

@@

The constraint that we impose in this case is the total number of opportunities $D$ in the region (compare to @eq-constraint0-gravitymodel):

@@

Substituting @eq-total-constrained-accessibility in @eq-total-constraint-accessibility, and solving for $k$, we have that (compare to the constant in the total flow spatial interaction model, Equation 2.11 in @cliff_evaluating_1974):

$$
k^D=\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
$${#eq-total-opportunity-balancing-factor}

Let us write our total opportunity constrained accessibility as follows:

$$
V^T_i = k^D\sum_j W^{(2)}_jf(c_{ij}) = \frac{D}{\sum_i\sum_j V^0_{ij}}\sum_j W^{(2)}_j f(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
$$

Further, we can see that, since $O$ and $W^{(2)}_j$ are opportunities, the following term is dimensionless:

@@

\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination:

$$
V^T_i = \sum_j W^{(2)}_j\kappa^D_{ij} 
$$

Notice that the role of $\kappa^0_{ij}$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region. Constant $\kappa^0_{ij}$ then assigns opportunities in proportion to the impedance between $i$ and $j$. For this reason we call it a proportional allocation factor.

## Total population constrained accessibility: market potential

One might wish to measure accessibility to the mass of the _population_ at $i$ instead of the opportunities at $j$. This is equivalent to the concept of market potential:

@@

$$
M_i^0 = W_i^{(1)} f(c_{ij})
$$

The population-constrained market potential becomes:

@@

The constraint that we impose in this case is the total market potential equals the total population $O$ in the region:

@@

## Singly-constrained accessibility: opportunity-constrained

If we decide to introduce additional information into our analysis, say, the number of opportunities _by destination_ (i.e., $D_j$), we can impose the following constraint (compare to @eq-constraint2-gravitymodel):

>>>>

---
title: "A family of accessibility measures derived from spatial interaction principles"
format: pdf
bibliography: "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
includes:
      in_header: 'gt_packages.sty'
      #\usepackage{float}
---

```{r knitr-setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  cache = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.pos = 'H', #to print figures on the line specified in knitted output
  comment = '', 
  out.width = "1\\linewidth")
```

@@

Of note, however is that between @stewartDemographicGravitationEvidence1948 and @hansen1959 the proportionality constant $G$ in @eq-stewart-population-potential vanished. There is some evidence that @hansen1959 was aware of the importance of this constant as he wrote about _directly_ and _inversely proportional_ relationships when discussing population, opportunities, and their separation in space. At any rate, those reading @hansen1959 must recall that @stewartDemographicGravitationEvidence1948 had set the proportionality constant $G$ to 1, with a note that "$G$ [was] left for future determination: a suitable choice of other units can reduce it to unity".

@@

Building on this, a family of spatial interaction models can be derived. In the framework introduced in @wilson1971, this includes the "unconstrained" (which we refer to as the total flow constrained) model in @eq-phys-gravity-model. The proportionality constant $k$ in this case is [see @cliff_evaluating_1974; @fotheringham_spatial_1984]:

@@

The work of @hansen1959 and @wilson1971 responded to important developments in the Anglo-American world at the time, in particular a need "to meet the dictates and needs of public policy for strategic land use and transportation planning" [@battyChronicleScientificPlanning1994]. Said dictates and needs were far from trivial. In the United States alone, the Federal-Aid Highway Act of 1956 authorized the creation of the U.S. Interstate Highway System, with a budget that ultimately exceeded one hundred billion dollars [@weinerUrbanTransportationPlanning2016; @mdotMnDOTJoins2007]. Spatial interaction modelling, with its ability to quantify trips, was incorporated into institutional modelling practices meant to "predict and provide", i.e., predict travel demand and supply transportation infrastructure [@kovatch1971modeling; @weinerUrbanTransportationPlanning2016]. Accessibility at the time did not quite have that power, as it did not quantify trips, but rather something somewhat more elusive: the less tangible "potential for interaction". In this way, where spatial interaction modelling became a key element of transportation planning practice, accessibility remained a somewhat more academic pursuit, and the two streams of literature only rarely connected.

@@

After compiling this corpus of documents, we used the {bibliometrix} package [@bibliometrix2017] to create a bibliographical coupling matrix. In this matrix, two documents are coupled if they share at least one common reference, and the strength of the coupling increases with the number of references that the documents have in common. @fig-Fig1 presents the results, where each symbol represents a document. The distance between symbols on the plot indicates the strength of bibliographical coupling: documents that are more strongly coupled (i.e., those sharing more references) are plotted closer to each other.

@@

```{r}
#| label: fig-Fig1
#| out-width: 70%
#| fig-cap: "Bibliometric coupling of papers that cite Hansen (1959), Wilson (1971), or both."

@@

As noted, literature that cites both @wilson1971 and @hansen1959 are sparse (only `r (n_b / (n_b+n_h+n_w) )|> scales::percent(0.1)` of the corpus) but highly coupled. However, we can discern from a manual reading of these works that they too are divergent, with one stream focused on developing accessibility (i.e., _potential for_ spatial interaction) and another on spatial interaction. The focuses of these divergent streams contribute this paper's broader hypothesis that the concepts of accessibility and spatial interaction have remained largely disconnected.

The spatial-interaction-focused stream of literature inspired by @wilson1971 that cites both @hansen1959 and @wilson1971, tends to contribute to understanding how accessibility is interpreted and incorporated in spatial interaction models, treating it as a separate (but related) phenomenon or variable. Specifically, works interpret the spatial interaction model's balancing factors (@eq-production-constrained-balancing-factor or @eq-doubly-constrained-balancing-factors) as the inverse of @hansen1959's model [@harrisEquilibriumValuesDynamics1978; @leonardiOptimumFacilityLocation1978; @fotheringhamSPATIALSTRUCTUREDISTANCE1981; @fotheringhamSpatialCompetitionAgglomeration1985], recognizing it as a "common sense" approach [@morrisAccessibilityIndicatorsTransport1979, p. 99] to including accessibility in the spatial interaction model, though further exploration of its relationship is warranted [@battyMethodResiduesUrban1976]. Some authors have explored this relationship, for instance as in @fotheringhamSpatialCompetitionAgglomeration1985 who demonstrates how the spatial interaction model may insufficiently explain spatial patterns and suggest that explicitly defining destinations' accessibility as a variable within the model may remedy the issue (e.g., the _competition destination_ model). Other works used both @hansen1959 and @wilson1971's framework in conjunction, such as in defining location-allocation problems in Operations research d[@leonardiOptimumFacilityLocation1978; @beaumontLocationallocationProblemsPlane1981], estimating trip flows (or some other interaction flows) alongside accessibility [e.g., @clarke2002deriving; @grengs2004measuring; @turk2019socio], or considering accessibility within spatial interaction models, in line with @fotheringhamSpatialCompetitionAgglomeration1985's demonstration [e.g., @beckers2022incorporating]. Other works departed from @hansen1959's definition and aligned with spatial interaction in different ways, such as using micro-economic consumer behaviour concepts to express potential for spatial interaction [@morrisAccessibilityIndicatorsTransport1979; @leonardiRandomUtilityDemand1984].

On the other hand, we discern that the potential-for-spatial-interaction-focused stream of literature inspired by @hansen1959 that cite both @hansen1959 and @wilson1971, cite @wilson1971 for three general reasons: 
- Firstly, as attribution for using using different context-dependent travel cost functions [@weibullNumericalMeasurementAccessibility1980; @handyMeasuringAccessibilityExploration1997; @kwan1998space; @shenLocationCharacteristicsInnercity1998;@ashiru2003space; @rau2012spatial; @pan2013impacts; @margaridacondecomelhoradoImpactMeasuringInternal2016; @caschili2015accessibility; @grengs2015nonwork; @pan2020measuring; @chia2020extending; @roblot2021participation; @sharifiasl2023incorporating; @kharel2024examining]. 
- Secondly, associating spatial interaction with accessibility's potential for spatial interaction [e.g., @giuliano2010accessibility; @grengs2010intermetropolitan; @grengs2010job; @grengs2012equity; @levine2012does; @levinson2012positive; @tong2015transportation; @liu2015spatial; @he2017measuring; @wuUnifyingAccess2020; @ng2022reflection; @naqavi2023mobility; @suel2024measuring]. While accessibility is an expression of _potential for_ spatial interaction and @wilson1971 touches on the concept [e.g., how @wilson1971 is cited in @millerMeasuringSpaceTimeAccessibility1999], it is distinct concept from spatial interaction. This distinction, however, is often unclear, as @hansen1959 and @wilson1971 are frequently co-cited as both being 'gravity models' [e.g., @liu2004accessibility; @dai2017visualization; @shen2019segregation; @chia2020extending].
- Thirdly, interpretation of @hansen1959's model as an inverse balancing factor [@vickermanAccessibilityAttractionPotential1974], especially in competitive opportunity or population contexts [@karstEvaluationAccessibilityImpacts2003;@geurs2006accessibility; @willigers2007accessibility;@el2011place; @curtis2010planning; @manaugh2012makes; @chen2013regional; @alonso2014labour; @albacete2017measuring; @sahebgharani2019computing; @mayaud2019future; @allenMeasureCompetitiveAccess2020; @levinsonGeneralTheoryAccess2020; @marwal2022literature; @su2023untangling]. However, as outlined in preceding sections, we argue the resulting indicator values are plagued by similar interpretability issues. Of the literature in this stream, only the work of @soukhovIntroducingSpatialAvailability2023 and @soukhovMultimodalSpatialAvailability2024 cite @wilson1971's use of balancing factors as ways to maintaining constraints in opportunities in the context of competitive accessibility.

@@

@@

As argued in the preceding section, the accessibility and spatial interaction modelling literature streams tended to evolve with little contact after @hansen1959 and @wilson1971. This may explain the failure of the constraints/proportionality constant(s) of spatial interaction models to cross over to accessibility analysis. This is intriguing since Wilson himself made an effort to connect his developments in spatial interaction modelling to accessibility, noting the denominator of @eq-production-constrained-balancing-factor and is the inverse of $A_i$, offering a potential interpretation for the proportionality constant [@wilson1971 p. 10]:

@@

This, however, only helps to illustrate the ease with which confusion may arise in this context. The relationship in @eq-Ai-as-accessibility appears to be misaligned. In the preceding section on early research, we noted that Hansen, and Stewart before him, defined accessibility as a partial sum of the demographic force $F$, essentially going from the demographic force ($F = G\frac{M_1\cdot M_2}{d_{12}^2}$ to the pairwise population potential $V_1 = G\frac{M_2}{d_{12}}$ and from there to the system-wide population potential (after losing $G$) $V_i = \sum_j \frac{M_j}{d_{ij}}$. While the inverse of $A_i$ may appear to be accessibility, its functional role in Wilson's general model $T_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})$ is that of a balancing factor $k$ (maintaining proportionality).

Hence, we propose a more useful definition: accessibility as a partial sum of the spatial interaction. A way to define the _potential for spatial interaction_ (i.e., accessibility) is as follows:

@@

In the following subsections, we detail how various accessibility measures can be defined depending on selected constraints: how $k$ is specified and depending on the combination of population and opportunity data is available and measurement focus. Together, these constraints delineate different members of the family of accessibility measures. Along with a simple numerical example, we detail the unconstrained case (@hansen1959's formulation), then total constrained cases (opportunity-constrained and population-constrained), then singly-constrained cases (opportunity-constrained and population-constrained), and lastly the doubly-constrained case. These cases are presented in increasing order of constraint restrictiveness and information availability.

## Simple numeric example setup

```{r}
#| label: simple-numerical-example-opportunities-and-population
id <- c("1", "2", "3")
O_i <- c(4, 10, 6) 
D_j <- c(160, 150, 180)
LU <- data.frame(id, O_i, D_j)
```

Consider a simple region with three zones, IDs 1, 2 and 3. Each zone is both an origin $i$ and a destination $j$. The following three pieces of information are defined: zonal population and opportunities, zonal cost matrix, and travel impedance functions for three types of travel behaviour. In this example, the population are people, the opportunities are physicians, and the region can be described by three possible travel behaviours.

Firstly, @tbl-small-system-land-use summarises the population (in units of 10,000s of people) and the opportunities (the number of physicians) per zone. ^[For reference, the number of physicians per 10,000 in Canada in 2022 was 24.97 [@whoMedicalDoctors102025]]. Considering @tbl-small-system-land-use's values, the Provider-to-Population-Ratio (PPR) in this system is `r sum(LU$D_j) / sum(LU$O_i)`. The population is $W_i^{(1)}$ when used as a proxy for the mass at the origin, and $O_i$ when used as a constraint. Similarly, the opportunities are represented by $W_j^{(2)}$ when used as a proxy for the mass at the destination, and $D_j$ when used as a constraint.

LU |>
  gt() |>
  # tab_spanner("{{W[_i^(1)]}}",
  #             columns = 2) |>
  # tab_spanner("{{W[_j^(2)]}}",
  #             columns = 3) |>
  cols_label(id = "ID (i or j)",
             O_i = "Population", 
             D_j = "Opportunities")  |>
  cols_align(align = "center",
             columns = 2:3) |>
  tab_footnote(footnote = md("Population is *Wi^(1)^* when used as a proxy for the mass at the origin, and *Oi* when used as a constraint."),
               locations = cells_column_labels(columns = O_i)) |>
  tab_footnote(footnote = md("Opportunities is *Wj^(2)^* when used as a proxy for the mass at the destination, and *Dj* when used as a constraint."),
               locations = cells_column_labels(columns = D_j)) |>
  as_latex()
```

Secondly, to pair with the zonal population and opportunity information, the assumed cost of movement (in minutes of travel time) between origins and destinations is as shown in @tbl-small-system-cost. From both @tbl-small-system-land-use and @tbl-small-system-cost, it can be interpreted that Zones 1 and 3 are more proximate to each other than to Zone 2. Combined, Zone 1 and 3 have the same amount of population as Zone 2 but more than double the opportunities. <!--for the readers, what sort of set-up could this reflect practically? -->

C <- expand.grid(oid = c("1", "2", "3"),
                  did = c("1", "2", "3")) |>
  mutate(cost = c(10, 30, 15, 30, 10, 25, 15, 25, 10)) |>
  mutate(f1 = cost^-3,
         f2 = cost^-2,
         f3 = cost^-0.1)
```

And lastly, we distinguish accessibility measure values for the following three impedance functions that represent the potential for spatial interaction travel behaviour of the population to opportunities in the region:
$$
\begin{array}{l}
f_1(c_{ij}) = \frac{1}{c_{ij}^3}\\
f_2(c_{ij}) = \frac{1}{c_{ij}^2}\\
f_3(c_{ij}) = \frac{1}{c_{ij}^{0.1}}
\end{array}
$$

Accessibility will be calculated three times for each case, one assuming the most decay ($f_1(c_{ij})$), another assuming medium decay ($f_2(c_{ij})$), and a third assuming the least decaying travel behaviour ($f_3(c_{ij})$) for the entire region. A helpful analogy may be tying travel behaviour to the used mode's mobility potential, i.e., the most decaying travel behaviour ($f_1(c_{ij})$) would assume all travel in the region being done by foot, while calculating accessibility assuming the least decay ($f_3(c_{ij})$) would assume unfettered automobility. Or alternatively, these functions could represent travel behaviour on snowstorm-affected day ($f_1(c_{ij})$) for the entire region versus a clear, ideal travel day ($f_3(c_{ij})$). As an example of a discussion on how travel behaviour has been considered in accessibility measures cost of travel see @paez2012measuring.

Any set of concepts representing population, opportunities, and their associated travel behaviour -whether representing the entire region uniformly (as we will demonstrate) or representing specific subgroups- can be substituted into our simple example, depending on the research question. The purpose of the following simple example is to demonstrate the calculation of each member of the accessibility measure family, interpret the values, and compare them both within and across travel behavior groups and members of the accessibility family.

@@

Setting the balancing factor $k$ to one in @eq-access-01 results in the unconstrained accessibility case:

@@

Then, the partial sum becomes the following: simply Hansen's accessibility $S_i$ [@hansen1959], standard practice in accessibility measurement:

@@

Where the sum of $V^0_{i}$ in the region generally does not equal the total number of opportunities $O$ (e.g., $\sum_i V^0_{i} \not= O$), as setting $k$ to 1 to arbitrarily eliminates the term and imposes no meaningful interpretation to the resulting values. To complicate matters, when comparing $V^0_{i}$ values between travel behaviour contexts, their difference depends on the choice of impedance function $f(c_{ij})$, associated parameters, and the number of zones for which an accessibility score is calculated. For this reason, the unconstrained accessibility should more appropriately be used as an ordinal variable to make comparisons of size (i.e., greater than, less than, equal to), not to calculate ratios or intervals (i.e., the magnitude of differences).

Returning to the simple example, consider the calculated unconstrained accessibility $V^0_{i}$ for each origin, a sum of all the travel impedance weighted opportunities at each destination ($\sum_i V^0_{i}$), in @tbl-simple-example-unconstrained-accessibility.

```{r}
#| label: tbl-simple-example-unconstrained-accessibility
#| tbl-cap: "Simple system: unconstrained accessibility."

unc_acc |>
  gt(rowname_col = "oid") |>
  tab_stubhead(label = "Zone") |>
  tab_spanner(label = "{{V[_i^0]}}",
              columns = 2:4,
              level = 2) |>
  tab_spanner(label = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
              columns = 2,
              level = 1) |>
  tab_spanner(label = "{{f_2 (c_ij ) = 1/c[_ij^2]}}",
              columns = 3,
              level = 1) |>
  tab_spanner(label = "{{f_3 (c_ij ) = 1/c[_ij^0.1]}}",
              columns = 4,
              level = 1) |>
  fmt_number(decimals = 3) |>
  cols_label(V_unc_i_1 = md("units: *f_1(c_ij)-weighted physicians"),
             V_unc_i_2 = md("units: *f_2(c_ij)-weighted physicians"),
             V_unc_i_3 = md("units: *f_3(c_ij)-weighted physicians"))  |>
  # cols_label(V_unc_i_1 = "{{f_1 (c_ij ) = 1/c[_ij^3]}}",
  #            V_unc_i_2 = "{{f_2 (c_ij )  = 1/c[_ij^2]}}<br>*physicians min^-2*",
  #            V_unc_i_3 = "{{f_3 (c_ij )  = 1/c[_ij^0.1]}}<br>*physicians min^-0.1*",
  #            .fn = md)  |>
  cols_align(align = "center",
             columns = 2:4) |>
  cols_width(
    stub() ~ px(60),
    everything() ~ px(180)
  ) |>
  tab_options(column_labels.font.size = "small") |>
  as_latex()
```

As the different impedance functions represent different travel behaviours, comparing unconstrained accessibility values across groups meaningless. For instance, at zone 1 the difference between the least decay ($f_3(c_{ij})$) and most decay ($f_1(c_{ij})$) groups is `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_1[1]`, but in what units? These two values, though representing the same zone but different groups in that zone, have different units so this difference value cannot be meaningfully interpreted.

However, we can compare values within the same travel behavior scenario across different zones but values are not inherently meaningful due to the units. For example, considering the most decaying scenario at zone 1, it captures `r unc_acc$V_unc_i_1[1] - unc_acc$V_unc_i_1[2]` more $f_1(c_{ij})$-weighted-physicians than zone 2, and `r unc_acc$V_unc_i_1[3] - unc_acc$V_unc_i_1[1]` fewer than zone 3. In the least decaying scenario, zone 1 captures `r unc_acc$V_unc_i_3[1] - unc_acc$V_unc_i_3[2]` more $f_3(c_{ij})$-weighted-physicians physicians than zone 2, and `r unc_acc$V_unc_i_3[3] - unc_acc$V_unc_i_3[1]` fewer than zone 3. Both scenarios at zone 1 captures an intermediate $V_i^0$ value between the two other zones, but at different ratios: the least decaying scenario captures `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[2]` and `r unc_acc$V_unc_i_1[1]/unc_acc$V_unc_i_1[3]` times the $f_1(c_{ij})$-weighted-physicians than zone 2 and 3 respectively, while the most decaying scenario captures `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[2]` and `r unc_acc$V_unc_i_3[1]/unc_acc$V_unc_i_3[3]` times the $f_3(c_{ij})$-weighted-physicians. The ratios between the zones in the two scenarios different. These differences are tied up in the travel impedance functions, and hence the unconstrained accessibility values themselves should not be compared. <!--maybe a better way of saying this?--> This simple example illustrates that the differences and ratios of unconstrained accessibility values are not inherently meaningful. While one could attempt to adjust the units post-calculation (e.g., scaling, population normalization) or select comparable impedance functions (at the cost of accurately reflecting travel behavior), such adjustments may introduce bias.

<!-- Notice the wild differences. Accessibility according to impedance function 3 basically says that everyone has access to all opportunities all at once. Accessibility according to function 1, on the other hand, says that each population has accessibility to around 3 opportunities, which is absurd: the number of opportunities available locally are in the order of 150-180. One should not have to adjust the impedance function to obtain "reasonable" results [EXAMPLE]; instead, the impedance function should reflect the travel behavior. -->

## Totally-constrained accessibility

A total constraint proportionally adjusts zonal accessibility values based on the total population and/or opportunities in the region. The introduction of this constraint introduces a form of competition, as all zonal values are a proportion of the system total, be it the regional opportunities or regional population depending on the case. We define two cases: a _totally-constrained opportunity_ accessibility case, interpreted as Hansen's accessibility with a constraint, and a _totally-constrained population_ accessibility case, interpreted as constrained market potential.

### Totally-constrained opportunity: Hansen's accessibility with a constraint

Unlike in @eq-access-01, the proportionality constant $k$ is retained and not set arbitrarily to 1:

@@

Where the total opportunity constrained accessibility measure now becomes Hansen's accessibility with a proportionality constant:

@@

The constraint that we impose in this case is the total number of opportunities $D$ in the region.

@@

This constraint is analogous to the total constraint of @eq-constraint0-gravitymodel which is congruent with Wilson's framework. We can then substitute @eq-total-constrained-accessibility in @eq-total-constraint-accessibility and solve for $k$ to yield: 
$$
k^D=\frac{D}{\sum_i\sum_j V^0_{ij}} = \frac{D}{\sum_i\sum_j W^{(2)}_jf(c_{ij})}
$${#eq-total-opportunity-balancing-factor}

Which is also congruent with Wilson's framework as it comparable to the total flow spatial interaction model (e.g., Equation 2.11 in @cliff_evaluating_1974). Hence, our totally-constrained opportunity accessibility model is:

$$
V^T_i = k^D\sum_j W^{(2)}_jf(c_{ij}) = \sum_j W^{(2)}_j\frac{D\cdot f(c_{ij})}{\sum_i\sum_j W^{(2)}_jf(c_{ij})} 
$$

Further, we can see that, since $D$ and $W^{(2)}_j$ are both in units of opportunities, the following term is dimensionless:

@@

\noindent and therefore $V^T_i$ is now in the units of $W^{(2)}_j$, that is, the mass at the destination ($V^T_i = \sum_j W^{(2)}_j\kappa^D_{ij}$). The role of $\kappa^D_{ij}$ in this reformulation of accessibility is to adjust the number of opportunities accessible from $i$ so that they represent a proportion of the total number of opportunities in the region. Constant $\kappa^D_{ij}$ then assigns opportunities in proportion to the impedance between $i$ and $j$. For this reason, we refer to it as a proportional allocation factor.

Referring to our simple numeric example, proportionality constant for $f_1(c_{ij}) = 1/c_{ij}^3$ would then be:
$$
\begin{array}{l}
k^D = \frac{D}{\sum_{i}\sum_{j} O_j f(c_{ij})}\\
k^D = \frac{D}{\frac{D_1}{c_{11}^3}+\frac{D_1}{c_{21}^3} + \frac{D_1}{c_{31}^3} + \cdots + \frac{D_3}{c_{31}^3} + \frac{D_3}{c_{32}^3} + \frac{D_3}{c_{33}^3}
}\\
k^D = \frac{490}{0.75} \\
k^D = 653.3256
\end{array}
$$

```{r}
# Calculate the balancing factors/proportionality constants for each of three impedance functions
k_tot <- unc_acc_ij |> 
  summarize(k1 = sum(LU$D_j)/sum(V_unc_ij_1), # f_1
            k2 = sum(LU$D_j)/sum(V_unc_ij_2), # f_2
            k3 = sum(LU$D_j)/sum(V_unc_ij_3)) # f_3
```

Using the calculated proportionality constants for all zones, the total opportunity constrained accessibility values for all zones and different travel behaviour scenarios is presented in @tbl-simple-example-total-opportunity-accessibility.

```{r}
#| label: tbl-simple-example-total-opportunity-accessibility
#| tbl-cap: "Simple system: total opportunity constrained accessibility."

In contrast to unconstrained accessibility, imposing a constraint -in this case a total opportunity constraint- allows for the comparison of differences and ratios. Each value is effectively in units of physicians, with the impedance units already accounted for by  $\kappa^D_{ij}$.

Consider the highest decay scenario ($f_1(c_{ij})$), zone 1 again captures an intermediate amount of physicians (`r tot_acc$V_tot_i_1[1]`). However, we can say that this is out of the `r tot_acc$V_tot_i_1 |> sum()` in the region, which allows us also to deduce that zone 1 captures `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[2]` and `r tot_acc$V_tot_i_1[1]/tot_acc$V_tot_i_1[3]` times more than zone 2 and 3. Values for the lesser decay ($f_2(c_{ij})$) and lowest decay ($f_3(c_{ij})$)  scenarios are calculated separately, with decay scenario values also summing to equal `r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region. Between zonal comparisons can be more interpretably made for these travel behaviour scenarios as well.

One can also directly compare values at a specific zone due to the consistent units. For instance, zone 1 remains intermediate in capturing accessible physicians relative to zones 2 and 3 across scenarios, similar to the unconstrained case. However, the difference between travel behaviour scenarios differ in direction. Specifically, Zone 1 captures `r tot_acc$V_tot_i_2[1]- tot_acc$V_tot_i_1[1]` more and `r tot_acc$V_tot_i_1[1]- tot_acc$V_tot_i_3[1]` fewer opportunities than the lesser decay scenarios $f_2(c_{ij})$ and $f_3(c_{ij})$ respectively. Why? $k^D$ ensures proportional allocation for each travel behaviour scenario. Meaning, while the unconstrained accessibility increases, $k^D$ adjusts the values to remain proportional to the total number of opportunities (`r tot_acc$V_tot_i_2 |> sum()` physicians accessible in the region). As the decay behaviour decreases, more opportunities are accessible for all zones. In the medium decay scenario $f_2(c_{ij})$, zone 1 sees a slight increase in values (relative to the highest decay scenario) as the zone can accessible more opportunities relative to increases seen in other zones. However, in the lowest decay scenario, zone 1 sees a decrease, as it is outpaced by increases in other zones - namely zone 2 (recall: zone 2 has the lowest number of opportunities, hence the increases in opportunity gains is much higher in a low decay scenario).

Using the total opportunity constrained formulation of accessibility offers a solution to the unit interpretability issue of @han1959's accessibility measure. Intuitively, the use of the constraint illustrates how the differences and ratios of values between zones and decay groups can be compared.

### Totally-constrained population: market potential

Effectively transposing $i$ and $j$, this case of the totally-constrained accessibility measure (@eq-total-constrained-market) is an expression of the concept of market potential (i.e., potential users) as proposed in @harris_market_1954 [@vickermanAccessibilityAttractionPotential1974]. The unconstrained form of market potential has been used in recent research  [e.g., @gutierrezLocationEconomicPotential2001; @holl2007twenty; @condecco2018road] to express the potentially accessible population (i.e., users) as a result of regional transportation infrastructure investment projects. To formulate this case, the total constraint may instead be applied to the mass of the _population_ at $i$ instead of the opportunities at $j$.

@@

$$
M_j^0 = W_i^{(1)} f(c_{ij})
$$

The population-constrained accessibility (market potential) becomes:

@@

The constraint that we impose in this case is the total market potential equals the total population $O$ in the region :

@@

Returning back to the numerical example, the proportionality constant would be solved for each travel behaviour scenario, and the market potential of each zone is expressed as units of population (e.g., the number of people accessible from each zone) in @tbl-simple-example-total-population-accessibility. Readers should not, zonals trends in accessible population and accessible physicians. in the preceding subsection (i.e., calculated using totally-constrained opportunity) are not the same: zone 1, 2, 3 represent destinations and the accessibility values represent the number of accessible people from the vantage of physicians.

<!--maybe we should say a little more? -->

```{r}
#| label: tbl-simple-example-total-population-accessibility
#| tbl-cap: "Simple system: total opportunity constrained accessibility."

## Singly-constrained accessibility

Increasing in restrictiveness relative to the totally-constrained accessibility measure, a single constraint proportionally adjusts the origin-destination zonal accessibility values based on the population or opportunities in the zone for all zones, while still maintaining the total system population or opportunity constraint, depending on the case. We define two cases: a _singly-constrained opportunity_ accessibility case and a _singly-constrained population_ accessibility case.

### Singly-constrained opportunity: spatial availability(?)
